The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens. Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps

### 2015-07-01 Manuscript/manuscript.tex
Discussion edits
Largely trying to emphasize differences between two scales of "trap effect"

This work demonstrates that (PCC: distribution may often be confused with spatial distribution rather than quality) habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population. (PCC: deleted, lots of keywords but no information).  In many cases, high quality habitat can enhance the disease spread and impact of environmentally persistent pathogens.  In addition, the habitats with many resources may become ecological traps that attract hosts, but are a net drain on the metapopulation due to the impacts of disease in those patches. This suggests a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability (PCC: rather than saying "processes", elaborate on what specific process you mean. ).

This work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and high pathogen endemicity (PCC: shouldn't there be a fourth scenario of pathogen drives the host extinct?). High quality habitats are a net benefit when pathogen spread is limited or is high enough that high and low quality patches are affected by disease. There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size. (PCC Note: most things you label as pathogen traits are actually emergent properties of the host-pathogen-environment...e.g. virulence)  (PCC: my text here makes some of the stuff below redundant)

\subsection*{Pathogens for which high quality habitat increases host population}

In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Figs~\ref{endemic}, \ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.

Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Figs~\ref{endemic}, \ref{highvlow}), high quality habitat is again a net benefit to the total host population size.  
Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to better maintain metapopulation stability in the face of widespread disease-induced mortality.

\subsection*{Pathogens for which high quality habitat decreases host population}

However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e. at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
In particular, the ability of these pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality. 
As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Fig~\ref{pinfection}).
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presence of an environmental reservoir or direct contact).

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat. (PCC: You're burying the lead. This stuff should go to the beginning of the discussion. )
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection.
By preventing the widespread disease-induced mortality observed in high quality habitat distributions, low quality distributions are then able to maintain larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).

Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of the trap effect (i.e. host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}). 
Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.  (PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.  Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts. This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)

@@

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a. to c.).  (PCC: would it be more accurate to describe the effect of envi. longevity as unimodal whereby at low longevities the pathogen works as though it were directly transmitted...and so there is a strong correlation between disease occupancy and host occupancy of a patch.  At intermediate longevity, low quality patches are important as they are less likely to be occupied allowing time for decontamination. At high levels of longevity, patches are not unoccupied for long enough for decontamination to occur.  If this is a fair representation, then we should avoid saying things like the 1st sentence as it implied a monotonic relationship. )
Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

The different scenarios and pathogens explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
The consistent patterns in the relative benefit of low and high quality habitat as a function of pathogen traits suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.  
By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g. plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
As we have seen, failing to consider the characteristics of a pathogen and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.

>>>>

This work demonstrates that habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
In many cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
In addition, the patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease. 
This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitats that hinder disease spread at the expense of metapopulation stability.

This expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and pandemic spread (possibly leading to host extinction). 
High quality habitats are a net benefit to the host population when pathogen spread is limited or is high enough that both high and low quality patches are affected by disease. 
In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.

There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size. 
In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.

Because high quality habitat supports larger host population sizes, these populations are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality. 
As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Supplement).
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presence of an environmental reservoir or direct contact).

Thus, if high quality habitat is preferred by migrants (i.e. $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Supplement) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, for pathogens with spread in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists. 
Thus, through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}, Supplement), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection.
By preventing the widespread disease-induced mortality observed in high quality habitat distributions, low quality distributions are then able to maintain larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).

Thus, for pathogens with this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
More importantly, they become ecological traps that expose migrant hosts to increased disease risk, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
Indeed, we see that the combination of host preference for high quality habitat, along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction.
From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Similarly, our model suggests that in the presence of some pathogens, maintaining or including low quality habitat in the metapopulation, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can facilitate larger total host populations than focusing on otherwise high quality habitat.  
(PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.  Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts. This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)).

@@

However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a metapopulation-level trap for a smaller range of potential pathogens (Fig~\ref{highvlow}, blue region shrinks from panels a. to c.).  (PCC: would it be more accurate to describe the effect of envi. longevity as unimodal whereby at low longevities the pathogen works as though it were directly transmitted...and so there is a strong correlation between disease occupancy and host occupancy of a patch.  At intermediate longevity, low quality patches are important as they are less likely to be occupied allowing time for decontamination. At high levels of longevity, patches are not unoccupied for long enough for decontamination to occur.  If this is a fair representation, then we should avoid saying things like the 1st sentence as it implied a monotonic relationship. )
Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.  
Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.  
These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g. from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g. increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g. decreased movement, low quality habitat).

The different scenarios and pathogens explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
The consistent patterns in the relative benefit of low and high quality habitat suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.  
By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g. plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
As we have seen, failing to consider the characteristics of a pathogen and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
